On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-21 manuscripts/PNAS/review/point_by_point_response.txt

We next respond to the reviewer's question about the mathematical "tightness" for the three types of quantities in our results : (a) required coding error eps(delta_1, delta_2) sufficient to guarantee uniqueness in terms of desired dictionary delta_1 and sparse representation error delta_2, (b) the constants C_1 and C_2 involved in our sufficient linear bound on eps(delta_1, delta_2) in terms of the noise, and (c) the number of sufficient samples N guaranteeing uniqueness.  With respect to (a), even with fixed dictionary, the recovery of sparse representations up to an error that is linear in the noise is essentially optimal (ACD13) (up to constants). Thus, the fact that we have a sufficient error in coding for uniqueness being linear for both the desired dictionary and sparse vector recovery error is optimal.  Although optimal, our solution to Problem 1 involves constants C_1 and C_2, and it would be interesting and relevant for practical applications to determine the minimal such constants. In the case of sparsity k=1, we have given an example showing optimality of constants in this case. Finally, with respect to (c), we regard this as a fundamental open problem in the field: how many sparsely generated samples determine the linear coding model?  (One reason to be skeptical that a polynomial suffice is again (TP14, T15).)  However, there is some good news -- if k is fixed or if the sparse codes \bar x_i are known to come from a polynomial number of supports, then a polynomial number of samples N is enough to guarantee uniqueness. We have revised the manuscript to discuss this last fact (which, incidentally, does not follow from the results of (HS15) in the noiseless case).

It is true, however, that we have not determined the tightest conditions regarding the constants and sample size necessary for uniqueness in sparse linear coding. However, we believe that these sufficient conditions yield strong enough conclusions to stand on their own; for instance, in synchronizing device parameters for a particular sparsely representable sensory ensemble.

### 2017-08-21 manuscripts/PNAS/review/point_by_point_response.txt

-----------------

RESPONSE:

We agree with the reviewer that a straightforward extension of the approach and noise-free result of (HS15) to the noisy case — while still a significant advancement in applied mathematics with actual practical implications — would not necessarily merit publication in PNAS on its own. However, our main result (Thm. 1) generalizes the noise-free result of (HS15) far beyond a simple extension to the noisy case in ways which provide valuable insight into the dictionary learning model and, moreover, a significant deviation from the approach taken in (HS15) was necessary to prove these new facts. Additionally, our solution (Cor. 2) to the optimization formulation of the dictionary learning problem (Prob. 2) — of most interest to optimization theorists and practitioners of their methods — is (to our knowledge) the first of its kind in both the noise-free and noisy domains. We discuss all of these points in more detail in our response to the reviewer’s later "detailed comments".

As per the reviewer’s request, we have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable). The main constants in our work -- C_1, C_2 -- are computable, elementarily, in terms of two positive quantities that are natural in our context: (a) the restricted "Lower Bound" L_k(M) of a matrix M, which is a measure of how distinguishable two (k/2)-sparse x’s are by their images under M, and (b) the value in Def. 4 of \xi(V_1,..,V_l) given subspaces V_i spanned by columns of A, which bounds the distance between intersections of such subspaces and each subspace individually. The first quantity (a) is, in some form or another, related to standard objects in the field and may be computed as the smallest singular value of a set of specific sub-matrices; in particular, for fixed sparsity constraint k, the lower bound L_k(A) is efficiently computable (polynomial time). Because deciding L_k(A) > 0 is already NP-hard (TP14), it seems likely that without fixing a sparsity level there is no efficient computation in general for (a). The calculation of (b) involves determining "canonical angles" between subspaces, which is usually reduced again to a numerical (efficient) singular value decomposition given the fixed subspaces. However, upon close inspection, C_2 nonetheless requires an exponential number of instances of computing \xi. Thus, while our constants are effective (an algorithm exists to compute them), they are not necessarily polynomial-time computable.
1
We next respond to the reviewer's question about the mathematical "tightness" for the three types of quantities in our results : (a) required coding error eps(delta_1, delta_2) sufficient to guarantee uniqueness in terms of desired dictionary delta_1 and sparse representation error delta_2, (b) the constants C_1 and C_2 involved in our sufficient linear bound on eps(delta_1, delta_2) in terms of the noise, and (c) the number of sufficient samples N guaranteeing uniqueness.  With respect to (a), even with fixed dictionary, the recovery of sparse representations up to an error that is linear in the noise is classically optimal (up to constants). Thus, the fact that we have a sufficient error in coding for uniqueness being linear for both the desired dictionary and sparse vector recovery error is optimal.  Although optimal, our solution to Problem 1 involves constants C_1 and C_2, and it would be interesting and relevant for practical applications to determine the minimal such constants. In the case of sparsity k=1, we have given an example showing optimality of constants in this case. Finally, with respect to (c), we regard this as a fundamental open problem in the field: how many sparsely generated samples determine the linear coding model?  (One reason to be skeptical that a polynomial suffice is again (TP14, T15).)  However, there is some good news -- if k is fixed or if the sparse codes \bar x_i are known to come from a polynomial number of supports, then a polynomial number of samples N is enough to guarantee uniqueness. We have revised the manuscript to discuss this last fact (which, incidentally, does not follow from the results of (HS15) in the noiseless case).

### 2017-08-21 manuscripts/PNAS/chazthm_pnas.tex
chaz

% TODO %
% Mention computability of eps(d1,d2) in terms of C1?
% *** Put ell1 norm into the denominator of L_k *** ?
% Give example of when we might expect k-uniformity, i.e. x_i in glp on k-dimensional subspaces. Orbit under group action of a k-dimensional subspace?

@@

If $\bar m < mr / (r-1)$ then, letting $p := \bar m - r( \bar m - m)$, we have $|T_1| = \bar m (r - 1) + p$ for positive $p$. By the pigeonhole principle again, there must then be at least $r$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is nonempty; hence, $|\cap \mathcal{G}_1| = 1$ by $r$-regularity and the SIP. Since $\cap \pi(\mathcal{G}_1)$ is nonempty, applying \eqref{cond} again gives $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [\bar m]$. If $p=1$, then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - r = (\bar m - 1)(r-1) + (p-1)$ ordered pairs having $\bar m - 1$ distinct first indices. Repeating the above arguments produces a (necessarily) distinct index $i_2$. Iterating the arguments $p$ times yields the set of singletons \mbox{$\bar J = \{i_1, \ldots, i_p\} \subseteq [\bar m]$}.
\end{proof}

>>>>

% TODO %
% bar J in Lem 4?
% Mention computability of eps(d1,d2) in terms of C1?
% *** Put ell1 norm into the denominator of L_k *** ?
% Give example of when we might expect k-uniformity, i.e. x_i in glp on k-dimensional subspaces. Orbit under group action of a k-dimensional subspace?

@@

If $\bar m < mr / (r-1)$ then, letting $p := \bar m - r( \bar m - m)$, we have $|T_1| = \bar m (r - 1) + p$ for positive $p$. By the pigeonhole principle again, there must then be at least $r$ elements of $T_1$ sharing the same first index. By \eqref{cond}, the intersection of the set $\mathcal{G}_1$ consisting of their second indices is nonempty; hence, $|\cap \mathcal{G}_1| = 1$ by $r$-regularity and the SIP. Since $\cap \pi(\mathcal{G}_1)$ is nonempty, applying \eqref{cond} again gives $\cap \pi(\mathcal{G}_1) = \{i_1\}$ for some $i_1 \in [\bar m]$. If $p=1$, then we are done. Otherwise, define $T_2 := T_1 \setminus \{(i,S) \in T_1: i = i_1\}$, which contains $|T_2| = |T_1| - r = (\bar m - 1)(r-1) + (p-1)$ ordered pairs having $\bar m - 1$ distinct first indices. Repeating the above arguments produces a (necessarily, by $r$-regularity and the SIP) distinct $\mathcal{G}_2 \subseteq \mathcal{H}$ and corresponding $i_2 \in [\bar m]$. Iterating the arguments $p$ times yields the set of singletons \mbox{$J = \{\cap \mathcal{G}_1, \ldots, \cap \mathcal{G}_p\} \subseteq [m]$}.
\end{proof}
